---
title: "Count Data, Poisson regression"
author: "Hana Akbarnejad"
date: "3/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(viridis)
library(ggplot2)
library(readr)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r data_import}

crab_data = read_table2("HW5-crab.txt") %>% 
  janitor::clean_names()

parasite_data = read_table2("HW5-parasite.txt") %>% 
  janitor::clean_names()
```

## Problem 1
### 1-a
Fit a Poisson model (M1) with log link with W(carapace width) as the single predictor. Check the goodness of fit and interpret your model.
```{r poi_w}

# fit poisson model
m1 = glm(sa ~ w, family=poisson(link = "log"), data=crab_data)
summary(m1)

# gof
deviance_m1 = m1$deviance
df_m1 =  m1$df.residual              # df=n-p  173-(1+1)=171
pval1=1-pchisq(deviance_m1,df=df_m1) # chisq test
pval1                                # pvalue is 0 this shows that the fit is not good
```

interpretation: .................
Assessing the goodness of fit of this model shows that the he p-value is `r round(pval1, 3)` which means that this model does not fit the data well.

### 1-b
Fit a model (M2) with W and Wt(weight) as predictors. Compare it with the model in (a). Interpret your results.
```{r poi_w_wt}

# fit poisson model
m2 = glm(sa ~ w+wt, family=poisson(link = "log"), data=crab_data)
summary(m2)

# compare m_2 and m_1
deviance_m2 = m2$deviance
df_m2 =  m2$df.residual             # df=n-p   173-(2+1)=170

test_stat =m1$deviance - m2$deviance
df =df_m1 - df_m2 
pval=1-pchisq(test_stat,df=df)      # chisq test
pval
```

interpretation: .................

Comparing the small model using only carapace width as predictor and the bigger model using both carapace width and weight as predictors, we can see that the p-value is `r round(pval, 3)`. This shows that we reject null hypothesis (stating that the smaller moddel is legible), and conclude that the bigger model is the one that we should pick and continue with.

### 1-c
Check over dispersion in M2. Interpret the model after adjusting for over dispersion.
```{r overdisp}

# calc dispersion parameter based on full model
p_res=residuals(m2,type='pearson',data=crab_data)
G=sum(p_res^2)
pval_disp=1-pchisq(G,df=170)
phi=G/170
phi
m2$deviance/m2$df.residual

summary(m2,dispersion=phi)

# # an equivalent way of estimating dispersion parameter
# wave.glm3 <- glm(damage~ship+year+period+offset(log(month)), family=quasi(link=log,variance=mu), data=wave, subset=month>0)
# summary(wave.glm3) # check ?quasi 
# 
# 
# # test over-dispersion (half normal plot)
# plot(qnorm((34+1:34+0.5)/(2*34+1.125)),sort(abs(res.p1)),xlab='Expected Half-Normal Order Stats',ylab='Ordered Abs Pearson Residuals')
# abline(a=0,b=1)
# abline(a=0,b=sqrt(phi),lty=2) 
```

interpretation: .................

The dispersion parameter is `r round(phi, 3)` for M2. After adjusting for over dispersion, it can be observed that the coefficients do not differ that much. However the covariate weight which was significant before adjusting for overdispersion becomes insignificant after adjusting for overdispersion.


## Problem 2
### 2-a
Fit a Poisson model with log link to the data with area, year, and length as predictors. Interpret each model parameter.
```{r}

mod1 = glm(intensity ~ area+year+length, family=poisson(link = "log"), data=parasite_data)
summary(mod1)
```

interpretation: .................

### 2-b
Test for goodness of fit of the model in (a) and state conclusions.
```{r mod1_gof}

deviance_mod1 = mod1$deviance
df_mod1 =  mod1$df.residual
pval_mod1=1-pchisq(deviance_mod1,df=df_mod1)
pval_mod1
```

The p-value is `r round(pval_mod1, 3)` this means that we reject the null and conclude that the model does not fit the data well.

### 2-c
Researchers suspect that there may be two strains of fish, one that is susceptible to parasites and one that is not. Without knowing which fish are susceptible, this could be regarded as a zero-inflated model. Building on the model in (a) (using the same predictors), fit an appropriate model to the data that can account for extra zeros. Provide an interpretation for
each model parameter in terms of the problem.
```{r}

```

