---
title: "Count Data, Poisson regression"
author: "Hana Akbarnejad"
date: "3/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(viridis)
library(ggplot2)
library(readr)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r data_import}

crab_data = read_table2("HW5-crab.txt") %>% 
  janitor::clean_names()
```

### 1-a
Fit a Poisson model (M1) with log link with W(carapace width) as the single predictor. Check the goodness of fit and interpret your model.
```{r poi_w}

# fit poisson model
m1 = glm(sa ~ w, family=poisson(link = "log"), data=crab_data)
summary(m1)

# gof
deviance_m1 = m1$deviance
df_m1 =  m1$df.residual              # df=n-p  173-(1+1)=171
pval1=1-pchisq(deviance_m1,df=df_m1) # chisq test
pval1                                # pvalue is 0 this shows that the fit is not good
```

interpretation: .................
Assessing the goodness of fit of this model shows that the he p-value is `r round(pval1, 3)` which means that this model does not fit the data well.

### 1-b
Fit a model (M2) with W and Wt(weight) as predictors. Compare it with the model in (a). Interpret your results.
```{r poi_w_wt}

# fit poisson model
m2 = glm(sa ~ w+wt, family=poisson(link = "log"), data=crab_data)
summary(m2)

# compare m_2 and m_1
deviance_m2 = m2$deviance
df_m2 =  m2$df.residual             # df=n-p   173-(2+1)=170

test_stat =m1$deviance - m2$deviance
df =df_m1 - df_m2 
pval=1-pchisq(test_stat,df=df)      # chisq test
pval
```

interpretation: .................

Comparing the small model using only carapace width as predictor and the bigger model using both carapace width and weight as predictors, we can see that the p-value is `r round(pval, 3)`. This shows that we reject null hypothesis (stating that the smaller moddel is legible), and conclude that the bigger model is the one that we should pick and continue with.

### 1-c
Check over dispersion in M2. Interpret the model after adjusting for over dispersion.
```{r overdisp}


```

