---
title: "Count Data, Poisson Regression"
author: "Hana Akbarnejad"
date: "3/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(viridis)
library(ggplot2)
library(readr)
library(pscl)
library(MASS)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r data_import, include=FALSE}

crab_data = read_table2("HW5-crab.txt") %>% 
  janitor::clean_names()

parasite_data = read_table2("HW5-parasite.txt") %>% 
  janitor::clean_names() %>% 
  mutate(
    area= as.factor(area),
    year= as.factor(year)
  )
```

## Problem 1
### 1-a
In this part, I want to fit a Poisson model (M1) with log link with W(carapace width) as the single predictor:
```{r poi_w}

# fit poisson model
m1 = glm(sa ~ w, family=poisson(link = "log"), data=crab_data)
m1_summary = summary(m1)
m1_summary
```

The model shows that the log rate ratio of the number of satellites residing near female crabs is `r round(m1_summary$coefficients[2], 3)` per one unit increase in female crab's carapace width. The p-value for this covariate is much smaller than 0.05 and shows this covariate is highly significant in the model. We can also see that the coeffcicient is positive which means that by increase in the width it is likely that on average there are more number of crabs with the female.

Now, I would like to Check the goodness of fit and interpret the model:
```{r}

# gof
deviance_m1 = m1$deviance
df_m1 =  m1$df.residual              # df=n-p  173-(1+1)=171
pval1=1-pchisq(deviance_m1,df=df_m1) # chisq test
pval1                                # pvalue is 0 this shows that the fit is not good
```

Assessing the goodness of fit of this model, we can see that the deviance is `r round(deviance_m1, 3)`. Comparing this value with $\chi^2$ with degree of freedom `r df_m1`, the p-value is `r round(pval1, 3)`. This means that this model does not fit the data well.

### 1-b
In this part I am going to fit a poisson model using both **weight** and **carapace width**:

```{r poi_w_wt}

# fit poisson model
m2 = glm(sa ~ w+wt, family=poisson(link = "log"), data=crab_data)
m2_summary = summary(m2)
m2_summary
```

The model shows that the log rate ratio of the number of satellites residing near female crabs is `r round(m2_summary$coefficients[2], 3)` per one unit increase in female crab's carapace width, holding their weight constant. The p-value for this covariate is `r round(m2_summary$coefficients[11], 3)` which shows this covariate is not significant in the model anymore.

Also, we can observe that the log rate ratio of the number of satellites residing near female crabs is `r round(m2_summary$coefficients[3], 3)` per one unit increase in female crab's weight, holding their carapace width constant. The p-value for this covariate is `r round(m2_summary$coefficients[12], 3)` which shows this covariate is significant in the model.

Next I want to compare the model I have fitted with the model that I had fit using only **carapace width** as covariate. We do not consider over dispersion in this part yet.

```{r m1_m2}

# compare m_2 and m_1
deviance_m2 = m2$deviance
df_m2 =  m2$df.residual             # df=n-p   173-3=170

test_stat =m1$deviance - m2$deviance
df =df_m1 - df_m2 
pval=1-pchisq(test_stat,df=df)      # chisq test
pval
```

Ignoring whether or not there is overdispersion in any of the models, when comparing the small model using only carapace width as predictor and the bigger model using both carapace width and weight as predictors, we can see that deviance is `r round(deviance_m2, 3)`. Comparing this devianve value with $\chi^2$ distribution with degree of freedom `r df` (the difference between the bigger model and df of smaller model), we observe that the p-value is `r round(pval,3)` This shows that we reject null hypothesis (stating that the smaller moddel is legible), and conclude that the bigger model is better, so we will consider **M2** as our model and move to the next part.

### 1-c
First I graph a half-normal plot to check if there is any over dispersion in model2.
```{r overdisp} 

# check overdispersion using half-normal plot
p_res=residuals(m2,type='pearson',data=crab_data)
G = sum(p_res^2)
phi=G/170
plot(qnorm((173+1:173+0.5)/(2*173+1.125)),sort(abs(p_res)),xlab='Expected Half-Normal Order Stats',ylab='Ordered Abs Pearson Residuals')
abline(a=0,b=1)
abline(a=0,b=sqrt(phi),lty=2)
```

The above plot shows that there is probably a source of overdispersion in the model because $\phi\neq 1$. Now, I will calculate the dispersion parameter and refit the model taking this parameter into account.

Equivalently, I could fit a negative binomial model to account for overdispersion without the need to calculate dispersion paarmeter and refitting the model.

I am going to do it both ways and show the results:

```{r}

# calc dispersion parameter based on full model
pval_disp=1-pchisq(G,df=170)
m2$deviance/m2$df.residual

m2_overdisp = summary(m2,dispersion=phi)
m2_overdisp

m2_nb=glm.nb(sa ~ w+wt,data=crab_data) ## glm.nb is a function from MASS to model negative binomial
summary(m2_nb)

```


The dispersion parameter is `r round(phi, 3)` for M2. After adjusting for over dispersion, it can be observed that the coefficients do not differ that much. However the covariate **weight** which was significant before adjusting for overdispersion becomes insignificant after adjusting for overdispersion. We can also see that the overalll results are similar using negative binomial model.

After adjusting for overdispersion, we can observe that the log rate ratio of the number of satellites residing near female crabs is `r round(m2_overdisp$coefficients[2], 3)` per one unit increase in female crab's carapace width, holding their weigt constant. The p-value for this covariate after adjusting for overdispersion is `r round(m2_overdisp$coefficients[11], 3)` which shows this covariate is not significant in the model.

Also, after adjusting for overdispersion we can observe that the log rate ratio of the number of satellites residing near female crabs is `r round(m2_overdisp$coefficients[3], 3)` per one unit increase in female crab's weight, holding their carapace width constant. The p-value for this covariate after adjusting for overdispersion is `r round(m2_overdisp$coefficients[12], 3)` which shows this covariate is not significant in the model.


## Problem 2
### 2-a
In this part, I fitted a Poisson model with log link function to the data with area, year, and length as predictors.
```{r}

mod1 = glm(intensity ~ area+year+length, family=poisson(link = "log"), data=parasite_data)
summary(mod1)
```

`r round(mod1$coefficients[2], 3)` is the log rate ratio of the number of parasites in area 2 versus area 1, holding year and length of the fish constant. 

`r round(mod1$coefficients[3], 3)` is the log rate ratio of the number of parasites in area 3 versus area 1, holding year and length of the fish constant. 

`r round(mod1$coefficients[4], 3)` is the log rate ratio of the number of parasites in area 4 versus area 1, holding year and length of the fish constant. 

`r round(mod1$coefficients[5], 3)` is the log rate ratio of the number of parasites in year 2000 versus 1999, holding area and length of the fish constant. 

`r round(mod1$coefficients[6], 3)` is the log rate ratio of the number of parasites in year 2001 versus 1999, holding area and length of the fish constant. 

`r round(mod1$coefficients[7], 3)` is the log rate ratio of the number of parasites per each one unit increase in the length of the fish, holding area and year constant.

Note that all these variables have p-value of smaller than 0.05, which shows these variables are significant in our model.

### 2-b
Now I am going to test the model I have built in previous part for goodness of fit.
```{r mod1_gof}

deviance_mod1 = mod1$deviance
df_mod1 =  mod1$df.residual
pval_mod1=1-pchisq(deviance_mod1,df=df_mod1)
```

To assess the good of fit of this model, I used deviance analysis and observed that the model has deviance of `r deviance_mod1` which is very high. Comparing this result with $\chi^2$ with degree of freedom of `r df_mod1` (63 observations deleted due to missingness), the p-value is `r round(pval_mod1, 3)`. This means that we should reject the null and conclude that the model does not fit the data well.

### 2-c
In this part I am interested in refit the model in part **a** that can account for extra zeros.

From `r nrow(parasite_data)` rows in dataset, there are `r parasite_data %>% filter(intensity == 0) %>% nrow()` observations with zero intensity of zero. We can consider two types of zeros in the context of this problem:

* True zeros: the strains which are not susceptible to parasites

* Pseudo zeros: the strains that are susceptible to parasites but have zero parasites int he time of testing, od the parasites have not been detected.So, I will fit a zero-inflated model.

I conditioned the model on area because the presence of parasite depends highly on the area we are investigating and I prefered to fix that when looking at the the difference between fish strains.
```{r zip}

mod2 = zeroinfl(intensity ~ year+length | area, data=parasite_data)
summary(mod2)
```

The model built has two parts: Count model and Zero-inflation model. 

**For the count model:**

Given that the fish strain is susceptible to parasites, `r round(mod2$coefficients$count[2], 3)` is the log rate ratio of the number of parasites in year 2000 versus 1999, holding length constant. 

Given that the fish strain is susceptible to parasites, `r round(mod2$coefficients$count[3], 3)` is the log rate ratio of the number of parasites in year 2001 versus 1999, holding length constant. 

Given that the fish strain is susceptible to parasites, `r round(mod2$coefficients$count[4], 3)` is the log rate ratio of the number of parasites per 1 unit increase in the length of the fish, holding year constant.

**For the zero-inflated model:**

`r round(mod2$coefficients$zero[2], 3)` is the log odds ratio of being the fish strain not being susceptible to parasites in area 2 compared to area 1.

`r round(mod2$coefficients$zero[3], 3)` is the log odds ratio of being the fish strain not being susceptible to parasites in area 3 compared to area 1.

`r round(mod2$coefficients$zero[4], 3)` is the log odds ratio of being the fish strain not being susceptible to parasites in area 4 compared to area 1.

(???)We can see that the rate of susceptible strain is higher in areas 2 and 3 compared to 1, but it is lower in area 4 compared to area 1. Also, all these variables are highly significant with p-values of smaller than 0.05. 
